---
title: "HW 4 - Turtles"
subtitle: "Part 4 - Drafting Viz"
description: "Due Tue 03/11/2025" 
author: "Eva Newby"
toc: true
warning: false
message: false 
eval: true
echo: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---


```{r}
# Load libraries
library(tidyverse)
library(here)
library(dplyr)
library(ggplot2)
library(sf)
library(ggspatial)
library(prettymapr)
library(rnaturalearth)
library(rnaturalearthdata)
library(forcats)
library(ggimage)
library(viridis)
library(viridisLite)
library(showtext)
library(geofacet)
```

### Read in Turtle Nesting Data and Font Data

```{r}
# Set the SHAPE_RESTORE_SHX config option
Sys.setenv("SHAPE_RESTORE_SHX" = "YES")

# Read the shapefile
shapefile_path <- "C:/MEDS/EDS240-dataviz/Assignments/newby-eds240-HW4/data/nestingbeaches_GCB.shp"
turtles <- st_read(shapefile_path) %>% 
    janitor::clean_names()

# Font options from google
font_add_google("Lora", "lora") 
font_add_google("Quicksand", "quicksand") 
font_add_google("Gochi Hand", "gochi")
showtext_auto()
```

### Exploratory Data Analysis and Data Wrangling

```{r}
# Check CRS
st_crs(turtles)

# Assume data into regional names
turtles <- turtles %>% 
    mutate(region = case_when(rmu == 1 ~ "Caribbean and Gulf of Mexico",
                              rmu == 2 ~ "Southwest Atlantic",
                              rmu == 3 ~ "Southeast Atlantic",
                              rmu == 4 ~ "Mediterranean",
                              rmu == 5 ~ "Arabian",
                              rmu == 8 ~ "Eastern Indian",
                              TRUE ~ "Unknown")) %>% 
    filter(region %in% c('Caribbean and Gulf of Mexico', 'Southeast Atlantic', 'Mediterranean', 'Arabian', 'Eastern Indian', 'Southwest Atlantic'))
```

### Data Viz 1 - Bar Chart
```{r}
# Summarize unique nest counts per region
region_counts <- turtles %>%
    st_drop_geometry() %>% 
    group_by(region) %>%
    summarize(nest_count = n()) %>%  # Count occurrences per region
    mutate(opacity_val = nest_count / max(nest_count))  # Normalize opacity (0-1)

# Join back to original data (so each region has its unique opacity)
turtles <- turtles %>%
    left_join(region_counts, by = "region")
```


```{r}
# Q2: World regions represented and their average nest count per year. ------------------------------------------------------------------------------- 
# per region

turtle_bar_plot <- turtles %>% 
    
    ggplot(aes(x = fct_rev(fct_infreq(region)),
               alpha = opacity_val))+
    
    geom_bar(fill = '#438D80')+
    
    geom_text(stat = 'count', 
              aes(label=after_stat(count)),
              hjust = 2,
              size = 8,
              color = "black",
              family = 'quicksand',
              fontface = "bold")+
    
    coord_flip()+
    
    geom_text(aes(x = region, y = 0, label = str_to_title(region)), 
             # family = 'gochi',
             family = "quicksand",
            #fontface = "bold",
            size = 7, 
            hjust = -0.03,
            nudge_y = 0.2) +
    
    labs(x = " ",
         y = "Number of Nests",
         title = " Average Loggerhead Turtle Nesting Counts ... (per year 1979 - 2023)")+
    
    theme_void()+
    
    theme(
        # text font
        text = element_text(family = "quicksand"),
        panel.grid.major.y = element_blank(),
        # axis.text.y = element_text(size = 10),
        plot.title = element_text(hjust = 0.5),

        # Sand colored background:
        panel.background = element_rect(fill = "#F4E7C5", color = NA),
        plot.background = element_rect(fill = "#F4E7C5", color = NA),
        legend.background = element_rect(fill = "#F4E7C5", color = NA),
        
        # Remove x and y axis labels
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
    ) +
    guides(alpha = "none")
        

turtle_bar_plot   
```

### Data Viz 2 - World Map
Countries in the caribbean/gulf of mexico  and Southeast Atlantic - HIGHLIGHT 
opacity value for nesting sites. general patterns by country
3 separate maps (top2 locations from bar chart)
or just caribbean

# Backup - what countries make up the caribbean/gulf of Mexico?


# This is the one!

```{r}
# Q2: Where are sea turtles nesting in the caribbean and gulf of Mexico ? ------------------------------------------------------------------

# Countried: Cayman Islands, USA, Mexico, Cabo Verde

# Filter to countries in the Caribbean/gulf of mexico and southeast Atlantic. 
map_turtles_filtered <- turtles %>% 
    filter(region %in% c("Caribbean and Gulf of Mexico", "Southeast Atlantic")) %>% 
    # Add a new column so each entry has an assigned bbox
    rowwise() %>% 
    mutate(bbox = list(bbox_list[[region]])) %>% 
    ungroup()

# Make Map -----------------------------------------------------------------------------------------
caribbean_map <- ggplot() +
    # add basemap
  geom_sf(data = world, fill = "#E6ECF1", color = "#E6ECF1") + 
    
    # add nesting layer
  geom_sf(data = map_turtles_filtered, fill = "#438D80", color = "#438D80") + 
    
    
    # set zoom to gulf of mexico and caribbean
  coord_sf(xlim = c(-97, -55), ylim = c(9, 33)) + 
    
  labs(title = "Turtle Sightings Around the Gulf of Mexico")+
    
    # apply a blank theme
  theme_void()

caribbean_map
```

```{r}
# Zoomed in map on USA
zoom_map_us <- ggplot() +
    # add basemap
  geom_sf(data = world, fill = "#E6ECF1", color = "#E6ECF1") + 
    
    # add nesting layer
  geom_sf(data = map_turtles_filtered, fill = "#438D80", color = "#438D80") + 
    
    
    # set zoom to gulf of mexico and caribbean
  coord_sf(xlim = c(-90, -75), ylim = c(25, 35)) + 
    
  labs(title = "Turtle Sightings Around the Southeastern USA")+
    
    # apply a blank theme
  theme_void()

zoom_map_us
```

```{r}
# Zoomed in map on Cabo Verde
zoom_map_cv <- ggplot() +
    # add basemap
  geom_sf(data = world, fill = "#E6ECF1", color = "#E6ECF1") + 
    
    # add nesting layer
  geom_sf(data = map_turtles_filtered, fill = "#438D80", color = "#438D80") + 
    
    
    # set zoom to cabo verde islands 
  coord_sf(xlim = c(-25, -20), ylim = c(14, 17)) + 
    
  labs(title = "Turtle Sightings in Cabo Verde")+
    
    # apply a blank theme
  theme_void()

zoom_map_cv
```

```{r}
# Zoomed in map on Eastern Mexico
zoom_map_mx <- ggplot() +
    # add basemap
  geom_sf(data = world, fill = "#E6ECF1", color = "#E6ECF1") + 
    
    # add nesting layer
  geom_sf(data = map_turtles_filtered, fill = "#438D80", color = "#438D80") + 
    
    
    # set zoom to cabo verde islands 
 coord_sf(xlim = c(-88, -87), ylim = c(20, 21)) + 
    
  labs(title = "Turtle Sightings in Eastern Mexico")+
    
    # apply a blank theme
  theme_void()

zoom_map_mx
```

```{r}
# Zoomed in map on Cayman Islands
zoom_map_ci <- ggplot() +
    # add basemap
  geom_sf(data = world, fill = "#E6ECF1", color = "#E6ECF1") + 
    
    # add nesting layer
  geom_sf(data = map_turtles_filtered, fill = "#438D80", color = "#438D80") + 
    
    
    # set zoom to caymen islands 
  coord_sf(xlim = c(-81.55, -81.15), ylim = c(19.15, 19.75)) + 
    
  labs(title = "Turtle Sightings in Caymen Islands")+
    
    # apply a blank theme
  theme_void()

zoom_map_ci
```



```{r}
# Q2: Where are sea turtles nesting in the caribbean and gulf of Mexico ? ------------------------------------------------------------------

# Set bounding boxes for region(s) of interest
bbox_carib <- st_bbox(c(xmin = -99, ymin = 9, xmax = -55, ymax = 35), crs = st_crs(turtles))

# Save as a dataframe
bbox_carib_df <- data.frame(xmin = -99, ymin = 9, xmax = -55, ymax = 35)
                     
# bbox_seat <- st_bbox(c(xmin = -50, ymin = 0, xmax = 0, ymax = 40), crs = st_crs(turtles))


# Save as list for easier access
bbox_list <- list(  "Caribbean and Gulf of Mexico" = bbox_carib,
                    "Southeast Atlantic"            = bbox_seat)

# Filter to countries in the Caribbean/gulf of mexico and southeast Atlantic. 
map_turtles_filtered <- turtles %>% 
    filter(region %in% c("Caribbean and Gulf of Mexico", "Southeast Atlantic")) %>% 
    # Add a new column so each entry has an assigned bbox
    rowwise() %>% 
    mutate(bbox = list(bbox_list[[region]])) %>% 
    ungroup()


# Get world map data
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter to caribbean region
caribbean_map <- world %>%
    st_crop(bbox_carib)

# Create the ggplot map for Caribbean and Gulf of Mexico
carib_map <- ggplot() +
    geom_sf(data = caribbean_map, fill = "gray") +
    geom_sf(data = map_turtles_filtered, color = "green", size = 8
            )
    theme_minimal()

carib_map
```


```{r}
# Where are sea turtles nesting in the gulf of mexico and caribbean?
map <-  ggplot(map_turtles_filtered) +

  # Initialize base rectangles and set opacity based on relative sighting proportions. *changed*----
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1, alpha = opacity_val), 
           fill = '#438D80') +
  
  # Add state abbreviations onto the rectangles ----
  geom_text(aes(x = 0.5, y = 0.7, label = country), 
            family = "quicksand", 
            fontface = "bold",
            size = 9, 
            color = "black") +
  
  # # Add number of observations per 10k residents onto the rectangles ----
  # geom_text(aes(x = 0.5, y = 0.3, label = num(pacity_value), 1)), 
  #           family = "quicksand", 
  #           fontface = "bold",
  #           size = 8, 
  #           color = "black") +  

  # Arrange rectangles in the USA mapping format by location ----
  geofacet::facet_geo(~country, grid = "world_countries_grid1") +
    
    
    # Add Caribbean bounding box
  geom_rect(data = bbox_carib_df, 
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            color = "black", fill = NA, size = 1.5)
    
    # empty theme
    theme_void() +
    
    # customize theme 
  theme(
    strip.text = element_blank(),  # Hide facet labels if needed
    plot.subtitle = element_text(family = "bar", size = 40, color = "#FF5733", hjust = 1, margin = margin(b = 10)),
    legend.position = "none"
  )

  # # Set aspect ratio to be fixed at an equal scale *changed*----
  # coord_fixed(ratio = 1) +
  # 
  # # Add a descriptive subtitle ----
  # labs(subtitle = "Sightings per 10k population") +
  # 
  # # apply a completely empty theme ----
  # theme_void() +
  # 
  # # further customize theme ----
  # theme(
  #   
  #   #  Remove state facet labels. *changed*----
  #   strip.text = element_blank(),
  #   
  #   # Customize subtitle size, color, and margin ----
  #   plot.subtitle = element_text(family = "bar", 
  #                                size = 40, 
  #                                color = accent,
  #                                hjust = 1,  
  #                                margin = margin(b = 10)),
  #   
  #   # Hide legend from view ----
  #   legend.position = "none"
    
  #)

map

```

### Data Viz 3 - Point and Line Chart
add in context/annotations
remove title

```{r}
# Add turtle icon
turtle_hatch <- "turtle-hatch.png"

# add turtle icon to each column that has a unique mann-kendall score value per country
turtles$image <- ifelse(duplicated(turtles$mann_kendal), NA, turtle_hatch)

# Filter for mann-kendall plot
mann_kendall_turtles = turtles %>% 
    filter(region %in% c('Caribbean and Gulf of Mexico'))

```

```{r}
# Just caribbean and gulf of mexico.
# Q3: Which regions that have nests have the highest variation in their mann-kendal slopes? ------------------------------------------------------

#  Mann-kendall plot per region

mann_kendall_plot <- mann_kendall_turtles %>%
    
  ggplot(aes(x = mann_kendal, 
             y = fct_rev(fct_infreq(region)))) +
    
  geom_line(aes(color = mann_kendal), 
            size = 4) +  
    
    # Use ggimage for turtle icons
  geom_image(aes(image = image), size = 0.06)+
    
    # Gradient color and gradient color legend
  scale_color_gradient2(low = "green", 
                        mid = "lightblue", 
                        high = "#438D80", 
                        midpoint = 0,
                        guide = guide_colorbar(title = NULL,
                                               barwidth = 23,
                                               barheight = 0.5),
                        breaks = c(-0.5, 0, 0.5),
                        labels = c("Decrease", "No Change", "Increase")) + 
    
    # "No change" h line
  geom_vline(xintercept = 0, 
             color = "black", 
             linetype = "dashed", 
             size = 0.8, 
             alpha = 0.5)+
  annotate("text", x = max(turtles$mann_kendal), 
           y = 0, 
           label = "No Change", 
           hjust = -1, 
           vjust = -30, # adjust this to move text
           angle = 90, 
           color = "black", 
           size = 3) +
    
    # label axes
  labs(x = "Change in Loggerhead Turtle Nesting Over Time (Mann-Kendall Slope)",
       y = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        # text font
        text = element_text(family = "quicksand"),
        
        # Sand colored background:
        panel.background = element_rect(fill = "#F4E7C5", color = NA),
        plot.background = element_rect(fill = "#F4E7C5", color = NA),
        legend.background = element_rect(fill = "#F4E7C5", color = NA),
        
       # plot.title = element_text(face = "bold", size = 14),
        axis.text = element_text(size = 12),
        legend.text = element_text(size = 10),
        
        # add dot/ turtle icon in the legend
        
        
        # increase plot margin on the right side
        plot.margin = margin(t = 10, r = 100, b = 10, l = 10)
    ) 

mann_kendall_plot
```
